Accelerating structure-function mapping using the ViVa webtool to mine natural variation

### 2018-12-06 manuscript/Supplement/ARFs.Rmd
bioRxiv submission

Auxin response is ultimately mediated by the auxin responsive transcription factors (ARFs). There are 23 ARFs in *Arabidopsis thaliana* that are divided into three phylogenetic classes. Class A ARFs (ARF5, ARF6, ARF7, ARF8 and ARF19) activate transcription. These ARFs have a glutamine-rich region in the middle of the protein that may mediate activation [@guilfoyle_auxin_2007]. It has recently been shown that the middle region of ARF5 interacts with the SWI/SNF chromatin remodeling ATPases BRAMA and SPLAYED, possibly to reduce nucleosome occupancy and allow for the recruitment of transcription machinery [@wu_auxinregulated_2015]. Additionally, ARF7 interacts with Mediator subunits, directly tethering transcriptional activation machinery to its binding sites in the chromosome [@ito_auxindependent_2016]. Class B and C ARFs are historically categorized as repressor ARFs, though the mechanism through which they confer repression has not been identified. Their middle regions tend to be proline- and serine-rich (citation, Ulmasov?).

@@

A few missense variants in *ARF5* highlighted potential functional variation in the class A ARFs. A high frequency allele *(What are the variants?)*, resulting in substitutions of small amino acids on the positive face of the PB1 domain for bulky phenylalanines, may disrupt interactions with other ARF or Aux/IAA PB1 domains. Interestingly, these variants were always found together in accessions Tu-B2-3, Ru-2, Haes-1, Rd-0, 627ME-13Y1, Le-0, Bu-0 and Bsch-0 **Where are these?**. Another highly represented polymorphism in ARF5 results in change in the 17th amino acid from a leucine to a methionine. This position is in the N-terminal extension of ARF5 which is of unknown function. It is possible that this methionine acts as an alternative start site in the accessions.

@@

ARF14, ARF15, **ARF16 I don't think this is correct ARF 16 is class C, perhaps this should be 12?**, ARF20, and ARF21 all do not have one or multiple of the conserved acidic residues in the PB1 domain, and all are within a subclade of the B ARFs. Interestingly, no non-synonymous variation is seen in these important residues for Aux/IAA and ARF interaction, suggesting that the amino acids that are in the regions may still be important for the function of these proteins. **Might cut this, it is interesting that perhaps homo/heterodimerization is not as important for this clade of ARFs but I think this conclusion mostly comes from looking at the alignments and not from the natural variation. Perhaps if it looks like these ARFs have higher than expected, or more deleterious variation across their PB1 domains**

@@

```{r ARF10_DBD Variants distribution}
C_ARFs <- subset(gene_info, tair_symbol %in% ARF_data[which(ARF_data$class == "C"), "name"])
C_ARFs <- alignCDS(C_ARFs$transcript_ID)
# make an alignment data frame
C_ARFs <- makeAlnDF(C_ARFs[[2]])
C_ARFs <- left_join(C_ARFs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
C_ARFs$tair_symbol %<>% factor(levels = ARF_order, 
                               ordered = TRUE)
C_ARFs$tair_symbol %<>% fct_rev()

@@

Several members of class B (ARF2, 3 and 4) and C ARFs (ARF10, 16 and 17) are distinguished from other ARFs on the basis of their post transcriptional regulation by trans-acting small interfering RNAs (tasiRNAs) and microRNAs (miRNAs) [@rhoades_prediction_2002; @allen_micrornadirected_2005]. **(AMY check for conservation of Arf 6 and 8 targeted by miRNA 167).** The miRNA target of class C ARFs showed a high degree of conservation (Fig. ARF10_16_17_Alignment). This is consistent with perturbation **what exactly does perturbation mean** studies that have shown drastic developmental phenotype for class C ARFs [@wang_control_2005; @mallory_micrornadirected_2005; @liu_repression_2007]. For the class B ARFs (2, 3 and 4) the tasiRNA binding motif is also highly conserved suggesting strong regulatory constraints on these genes (Fig ARF2_3_4_Alignment). The DNA binding domain of ARF4 exhibits non-synonymous variation in 4 accession and would be interesting to carry out phenotypic characterization of these accessions for developmental growth defects (Fig ARF4_DBD Variants map).

```{r ARF10_16_17_Alignment}

```{r ARF2_3_4_Alignment}

```{r ARF4_DBD Variants map}

>>>>

Auxin response is ultimately mediated by the auxin responsive transcription factors (ARFs). There are 23 ARFs in *Arabidopsis thaliana* that are divided into three phylogenetic classes. Class A ARFs (ARF5, ARF6, ARF7, ARF8 and ARF19) activate transcription. These ARFs have a glutamine-rich region in the middle of the protein that may mediate activation [@guilfoyle_auxin_2007]. It has recently been shown that the middle region of ARF5 interacts with the SWI/SNF chromatin remodeling ATPases BRAMA and SPLAYED, possibly to reduce nucleosome occupancy and allow for the recruitment of transcription machinery [@wu_auxinregulated_2015]. Additionally, ARF7 interacts with Mediator subunits, directly tethering transcriptional activation machinery to its binding sites in the chromosome [@ito_auxindependent_2016]. Class B and C ARFs are historically categorized as repressor ARFs, though the mechanism through which they confer repression has not been identified. Their middle regions tend to be proline- and serine-rich.

@@

@@

@@

```{r C-ARFs-alignment}
C_ARFs <- subset(gene_info, tair_symbol %in% ARF_data[which(ARF_data$class == "C"), "name"])
C_ARFs <- alignCDS(C_ARFs$transcript_ID)
# make an alignment data frame
C_ARFs <- makeAlnDF(C_ARFs[[2]])
C_ARFs <- left_join(C_ARFs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
C_ARFs$tair_symbol %<>% factor(levels = ARF_order, 
                               ordered = TRUE)
C_ARFs$tair_symbol %<>% fct_rev()

@@

### 2018-12-06 manuscript/Supplement/IAAs.Rmd
bioRxiv submission

The Aux/IAA proteins (IAAs) mediate both auxin perception and transcriptional co-repression of auxin responsive genes. Auxin perception is carried out along with TIR1/AFB proteins, which together with IAAs act as co-receptors to bind and perceive nuclear auxin. Interaction of IAAs with auxin and TIR1/AFBs is conferred by the highly conserved 13 amino acid degron motif [@tan_mechanism_2007]. Formation of this auxin co-receptor complex triggers ubiquitination and subsequent degradation of IAAs. This degradation in turn relieves their repression upon ARF transcription factors. Stabilizing mutations within the degron can lead to dominant phenotypes [@berleth_auxin_2004; @yang_iaa1_2004]. The *Arabidopsis thaliana* Aux/IAA family has 29 members, most of which possess a canonical auxin degron. Notable exceptions are the auxin-insensitive IAA20 and IAA30 that lack a degron sequence entirely, while the closely related IAA31 retains a semi-functional degron sequence with a very slow degradation rate [@dreher_arabidopsis_2006; @havens_synthetic_2012]. Additionally, IAA degradation rate is influenced by rate motifs found outside the degron, such as the KR motif [@dreher_arabidopsis_2006; @moss_rate_2015]. IAA-mediated repression of ARFs depends upon both an EAR domain, N-terminal of the degron, that recruits co-repressors of the TOPLESS (TPL) family, and a Phox and Bem1 (PB1) domain that enables interaction via hetero-oligomerization with ARF PB1 domains [@nanao_structural_2014; @kagale_ear_2011].

@@

###Sister Pair Diversification

Previous research has found evidence of both broad genetic redundancy and specificity within closely related pairs or groups of IAA proteins [@overvoorde_functional_2005]. For example, the *iaa8-1 iaa9-1* double mutant and the *iaa5-1 iaa6-1 iaa19-1* triple mutant have wild-type phenotypes [@overvoorde_functional_2005], yet the *IAA6/19* sister pair has significant differences in expression patterns, protein abundances and functions suggesting they have undergone functional specialization since their divergence [@winkler_variation_2017].

In our analysis of the *IAA* gene family, we frequently observed that one member of most *IAA* sister pairs had high nonsynonymous diversity while the other sister was more conserved \@ref(fig:IAAtree). In an interspecific comparison between *A. thaliana* and *A. lyrata*, *IAA19* is more conserved, while *IAA6* has regions of increased sequence divergence upstream of the degron and in the N-terminal half of the PB1 domain [@winkler_variation_2017]. Consistent with these results, our analysis of intraspecific natural variation revealed higher conservation for *IAA19* (${\pi_N/\pi_S}$ = 0.55) compared to *IAA6* (${\pi_N/\pi_S}$ = 2.3) \@ref(fig:IAAtree), and also detected high diversity within the same regions of *IAA6* as seen in Winkler et al. (Supp Fig IAA6 diversity plot).

@@

IAA_data <- IAA_data[match(gsub(x = tree@phylo$tip.label, pattern = "\\.1",replacement = ""), IAA_data$AGI), ] 
tree@phylo$tip.label <- IAA_data$name
```
**I think IAA20, 30, and 15 should also be in class C. So there are really two large clades A and B and then there are truncations/loss of function/dominant variants scattered throughout, Class C.**

(ref:IAAtree) **IAA protein sequence tree mapped with ${\pi_N/\pi_S}$.** Protein sequences were aligned [@wright_decipher_2015] and low information content regions were masked [@kuck_parametric_2010] prior to infering a phylogeny [@ronquist_mrbayes_2003]. Tips of the tree are mapped with circles of diameter proportional to ${\pi_N/\pi_S}$ and also are colored according to ${\pi_N/\pi_S}$. Nodes are labeled with the poster probability of monophyly.

```{r IAAtree, fig.height=5, fig.width=5, fig.cap="(ref:IAAtree)"}
p <- ggtree(tree, layout = "rectangular", color = "grey") %<+% IAA_data + theme_tree() + scale_x_continuous(expand = c(0, 1.2)) + 
  geom_tippoint(mapping = aes(size = 4*PiN_PiS, fill = PiN_PiS), shape = 21, color = "white", show.legend = TRUE) + 
  scale_fill_viridis(values = rescale(c(min(IAA_data$PiN_PiS), 1,
                                        max(IAA_data$PiN_PiS)), 
                                      to = c(0,1)), guide = "colourbar") + scale_size(guide = "none") + 
  theme(legend.position = c(.9,.7)) + 
  geom_nodelab(mapping = aes(label = round(as.numeric(prob),digits = 2)), nudge_x = -.3, nudge_y = .5, size = 3) + 
  geom_tiplab(mapping = aes(color = Class), hjust = -0.25) + 
  scale_color_viridis(discrete = TRUE)
p
```

Within the Group B *IAA* genes, there are four sister pairs---*IAA10*/*IAA11*, *IAA12*/*IAA13*, *IAA18*/*IAA26*, and *IAA20*/*IAA30* \@ref(fig:IAAtree) [@remington_contrasting_2004]. Interestingly, among these sister pairs only *IAA12* and *IAA13* have a noticeable difference in nonsynonymous diversity. The *Arabidopsis thaliana* EFP browser shows that *IAA10* and *IAA11* have almost identical expression patterns, with the exception of *IAA10* being more highly expressed in the developing seed, and *IAA11* being more highly expressed in root tissue during lateral root development **need to cite original data source here**. In our analysis, *IAA10* and *IAA11* both showed functional conservation (${\pi_N/\pi_S}$ of 0.80 and 0.67 respectively) in support of *IAA10* and *IAA11* playing redundant roles. Accessions and associated variants identified here provide a potentially expedited route to elucidating the extent of redundancy between family members such as *IAA10* and *IAA11*.

### New Loss-of-Function/Gain-of-Function alleles

Accessions with functional variants can be potentially used to investigate the extent of redundancy/specialization within sister pairs. For example, three accessions have a missense variant affecting the EAR domain of IAA10 (HR-5, UKSE06-118 and UKSW06-240, all from southern UK). This variant (L49S) is in the second leucine of the LxLxL motif and likely reduces the ability of IAA10 to recruit the TOPLESS co-repressor [@lee_biological_2016]. The nonsense variant (IAA10-Y219\*) resulting in truncated PB1 domain may be more indicative of *IAA10* null phenotype, as it likely prevents IAA10 from interacting with an ARF. This variant is only found in one accession from northeastern Spain (IP-Svi-0). These accessions and potential loss-of-function alleles may facilitate exploration of the function of *IAA10*.

Within the available natural sequence variation, we found several missense polymorphisms affecting the EAR domain or degron of several canonical clade A IAAs. However, no missense polymorphisms were present in the KR motif or residues critical for PB1-PB1 domain interactions (\@ref(fig:A-IAA-alignment)). The majority of missense mutations observed were in the variable regions between the conserved domains.

(ref:A-IAA-alignment) 
```{r A-IAA-alignment, fig.cap="(ref:A-IAA-alignment)"}
A_IAAs <- subset(gene_info, tair_symbol %in% IAA_data[which(IAA_data$Class == "A"), "name"])
A_IAAs <- alignCDS(A_IAAs$transcript_ID)
# make an alignment data frame
A_IAAs <- makeAlnDF(A_IAAs[[2]])
A_IAAs <- left_join(A_IAAs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
A_IAAs$tair_symbol %<>% as.factor()
A_IAAs$tair_symbol %<>% fct_reorder(
    {as.integer(str_remove(A_IAAs$tair_symbol, 
                           pattern = "IAA"))}, 
    .desc = TRUE)
VCF <- ldply(.data = VCF, .fun = subset, gt_GT != "0|0")
coding_VCF <- getCodingDiv(VCF)
A_IAAs <- addSNPsToAlnDF(aln_df = A_IAAs,SNPs = coding_VCF, effect_order = SNPeff_order)
A_IAAs$gap <- A_IAAs$seq_pos == "-"

p <-ggplot(A_IAAs, aes(x = aln_pos, y = tair_symbol)) +
  geom_raster(data = A_IAAs, 
             # mapping = aes(fill = effects)) +
              mapping = aes(fill = strength,
                            alpha = gap)) +
  scale_alpha_manual(values = c("TRUE" = 0, "FALSE" = 1), 
                     labels = c("aligned", "gap"), 
                     guide = guide_legend(override.aes = 
                                            list(fill = c("grey85", "white")),
                       title = "alignment",
                       direction = "vertical")) + 
   scale_fill_viridis(breaks = range(SNPeff_order$strength),
                      limits = range(SNPeff_order$strength),
                      na.value = "grey85",
                      direction = -1,
                      labels = c("neutral", "deleterious"),
                      option = "A", 
                      guide = guide_colorbar(ticks = FALSE, 
                                             title.vjust = .8, 
                                             title.position = "top", 
                                             title.hjust = 0.5)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +  # expand increases distance from axis
  labs(x = "codon position in alignment", y = "", fill = "variant effect") +
  theme_logo(base_family = "Helvetica") +
  theme(panel.grid = element_blank(), 
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        #panel.background = element_rect(fill = "grey85", color = "white"),
        legend.position = c("bottom"),
        legend.key = element_rect(linetype = "solid"),
        legend.spacing = unit(4, "char"), 
        legend.box.spacing = unit(0.1,"line")) 
p

ggsave("A_IAAs.png", width = 7, height = 4)
```

Several examples of possible gain-of-function missense variants were found in the degron of *IAA2*, *IAA3*, *IAA6*, *IAA7*, *IAA8*, *IAA13*, *IAA15*, *IAA18*, *IAA19*, *IAA27*, *IAA28* and *IAA29*. Most degron variants were very minor alleles in the population, only represented in one or two accessions. Interestingly, degron missense variants occured more frequently in *IAA* genes that have a closely related sister, which may indicate redundant function within the sister-pairs. 
**I think this is an interesting finding, but I'm not sure about this hypothesis of redundant function. If pairs are acting redundantly and we assume the degron variants are dominant, we would expect the auxin signal they transmit (their function) to be lower, in sum, as a pair. So I would actually hypothesize that there has been some divergence in function as a pair.** 
```{r Correlation between Distance to MRCA and prevalence of degron variants}

```{r Correlation between Pi_N/Pi_S and k_5}

*IAA6* has two potentially dominant missense mutations found within the degron (Supp Fig. alignment), one previously believed to lead to the *shy1-1* phenotype [@kim_two_1996]. Interestingly, this variant is widely represented in 31 accessions across Europe (Supplemental Map Fig). This finding supports the proposition that iaa6-C78R may not be the mutation causing the *shy1-1D* phenotype but rather represents natural variation in the degron sequence ([note from Mike Prigge in TAIR](https://www.arabidopsis.org/servlets/TairObject?type=notepad&id=11247)). These accessions could be used to confirm whether *iaa6-C78R* is responsible for the *shy1-1D* phenotype. Additionally, an accession was identified with the IAA6-V77M variant, also in the degron, that may be useful to further explore *iaa6* gain-of-function phenotypes.

### Identifying novel functional sequences in IAAs

An exciting possible use for the Natural Variation Webtool is to search for conserved, low-diversity sites that may represent possible sites of functional import, such as post-translational modification or protein-protein interaction. For example, recent detailed biochemical analysis identified putative lysine ubiquitination sites in members of the *IAA6/19* sister pairs [@winkler_variation_2017]; Supp Fig IAA6/19 alignment). These sites are conserved across natural sequence variation, while several other lysine residues vary (Supp Figs alignment and map). In IAA6, there are 17 lysines of which ~6 are putative ubiquitination sites~ **I think this should be 5, K3, K32, K33, K91, and K97. Missense variants occured in K34Q, K140N *. One is proximal to the KR motif, while two others show natural variation. In IAA19, there are 15 total lysines, 8 of which are putative ubiquitination sites (K3, K25, K68, L87, K93, K100, K111, and K141). Two of these ubiquitination sites are proximal to the KR motif and degron. **This sentence is not clear. Only 1 of the ubiquitination sites varies? or one lysine, I think you mean this only 1 lysine that varies, but it is not ubiquitinated.** Natural sequence diversity may be used to narrow down a subset of lysines to assess for functional ubiquitination.

```{r IAA6/19-alignment, fig.cap="(ref:IAA6/19-alignment"}

```{r IAA variants mapped to IAA17, eval=FALSE}
coding_vcf <- ldply(.data = VCF, .fun = subset, 
             !is.na(Transcript_ID) & gt_GT != "0|0")
aln_df <- addSNPsToAlnDF(aln_df, coding_vcf, seq_name = Transcript_ID, 
               seq_pos = Codon_Number)

selections <- select_variants(gene_info, aln_df, structure_ref = "IAA17")
selections$V2 <- paste0("chain X and resi ", selections$V2)
write.table(selections, file = "IAA_variant_selections", row.names = FALSE, 
          col.names = FALSE, sep = ",", quote = FALSE)
```

The members of the *IAA8/9/27* clade have a unique N-terminal extension with a highly conserved region common to orthologs in several other species [@dreher_arabidopsis_2006]. *IAA8* and *IAA9* play redundant roles in leaf serration [@koenig_auxin_2009], while *IAA8* also plays an important role in lateral root development [@arase_iaa8_2012]. IAA8 has been reported to uniquely localize in the cytosol [@arase_iaa8_2012] where it may interact with LSD1 protein [@coll_programmed_2011]. The N-terminal extension has low natural sequence diversity in *IAA8* and *IAA9* (Supp Fig IAA8/9 alignment) that may represent a site of novel functionality, possibly mediating interaction with LSD1. The few accessions with missense alleles in this region in both *IAA8* and *IAA9* may be valuable in probing the functionality of this region (Supplemental Fig IAA8/9 map).

```{r IAA8_9_27 alignment}

```{r IAA8_9 map}

Members of the *IAA17/7/14/16* clade have a second EAR motif (LxLxL) with no observed missense variants in *IAA7*, *IAA16*, or *IAA17* (Supp Fig IAA7/14/16/17 alignment). This observation supports the experimental evidence for IAA7 that the second EAR motif is functional, *i.e.* it can interact with TPR1 co-repressor and has a minor repressive role in auxin-related developmental phenotypes [lee_biological_2016]. For *IAA14*, there were two missense variants that may be of interest to probe functionality of this second EAR motif. One variant, L41F, in the central lysine is represented in only one accession, and the other in the second “x” (K40N) is found in twelve accessions around Copenhagen (Supp Fig. IAA14 map).

```{r IAA7_14_16_17 alignment}

```{r IAA14 map}

>>>>

@@

@@

IAA_data <- IAA_data[match(gsub(x = tree@phylo$tip.label, pattern = "\\.1",replacement = ""), IAA_data$AGI), ] 
tree@phylo$tip.label <- IAA_data$name
```

(ref:full-IAA-alignment) 
```{r full-IAA-alignment, fig.cap="(ref:full-IAA-alignment)"}
IAAs <- alignCDS(gene_info$transcript_ID)
# make an alignment data frame
IAAs <- makeAlnDF(IAAs[[2]])
IAAs <- left_join(IAAs, dplyr::select(gene_info, "tair_locus",
                                       "tair_symbol", "transcript_ID"),
                        by = c("transcript_ID" = "transcript_ID"))
IAAs$tair_symbol %<>% as.factor()
IAAs$tair_symbol %<>% fct_reorder(
    {as.integer(str_remove(IAAs$tair_symbol, 
                           pattern = "IAA"))}, 
    .desc = TRUE)
VCF <- ldply(.data = VCF, .fun = subset, gt_GT != "0|0")
coding_VCF <- getCodingDiv(VCF)
IAAs <- addSNPsToAlnDF(aln_df = IAAs,SNPs = coding_VCF, effect_order = SNPeff_order)
IAAs$gap <- IAAs$seq_pos == "-"

p <-ggplot(IAAs, aes(x = aln_pos, y = tair_symbol)) +
  geom_raster(data = IAAs, 
             # mapping = aes(fill = effects)) +
              mapping = aes(fill = strength,
                            alpha = gap)) +
  scale_alpha_manual(values = c("TRUE" = 0, "FALSE" = 1), 
                     labels = c("aligned", "gap"), 
                     guide = guide_legend(override.aes = 
                                            list(fill = c("grey85", "white")),
                       title = "alignment",
                       direction = "vertical")) + 
   scale_fill_viridis(breaks = range(SNPeff_order$strength),
                      limits = range(SNPeff_order$strength),
                      na.value = "grey85",
                      direction = -1,
                      labels = c("neutral", "deleterious"),
                      option = "A", 
                      guide = guide_colorbar(ticks = FALSE, 
                                             title.vjust = .8, 
                                             title.position = "top", 
                                             title.hjust = 0.5)) +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +  # expand increases distance from axis
  labs(x = "codon position in alignment", y = "", fill = "variant effect") +
  theme_logo(base_family = "Helvetica") +
  theme(panel.grid = element_blank(), 
        panel.grid.minor = element_blank(),
        strip.background = element_blank(),
        strip.text.x = element_blank(),
        #panel.background = element_rect(fill = "grey85", color = "white"),
        legend.position = c("bottom"),
        legend.key = element_rect(linetype = "solid"),
        legend.spacing = unit(4, "char"), 
        legend.box.spacing = unit(0.1,"line")) 
p

#ggsave("IAAs.png", width = 7, height = 4)
```

### 2018-12-06 manuscript/Supplement/TPLs.Rmd
bioRxiv submission

The TOPLESS co-repressor family generally exhibits a high level of sequence conservation at the amino acid sequence level across resequenced *Arabidopsis thaliana* accessions, with all ${\pi_N/\pi_S}$ values below 1 (Table X). The closely related *TPL* and *TPR1* have the highest ${\pi_N/\pi_S}$ values (TPL-0.425, TPR1-0.739), suggesting that these these two related genes tolerate a higher degree of sequence and potentially functional diversity compared to *TPR2/3/4*. The N-terminal TPL domain of the TPL/TPR family is particularly conserved (Figure X). All nonsynonymous polymorphisms observed in this region are either in the coils between helices or are highly conservative mutations within helices (i.e. Valine to Isoleucine), which would be predicted to exhibit little effect on folding and function.

@@

Compared to the TPL domain, both the proline rich and WD-40 domains exhibited a higher frequency of missense variants. We utilized Phyre2.0 modeling software to predict the structure of the TPL/TPR WD-40 repeats in order to map non-synonymous mutations to protein structure (Figure X). Here we can see both faces of each predicted beta-propeller structure of the WD-40 domains, which are connected by a short peptide linker. This model reveals amino acid conservation in the core and central residues of the WD-40 domains. ~~These observations could be used as the basis for forward mutagenesis strategies to test whether this region of TPL functions similarly to its homolog TLE, which binds its target genes EH1 and the C-terminal WRPW/Y motif (Hairy/Hes/Runx) in the center of the WD-40 domain (pmid 16762837).~~ **Can we say something more general here? There is not enough background to support this. What is TLE from? What is EH1? Is

@@

(ref:full-TPLs-alignment) **Alignment of A. thaliana TPL/TPR genes showing variants.** This alignment should probably be paired down to a key region rather than the eniter AA sequence.

```{r full-TPLs-alignment, fig.cap= "(ref:full-TPLs-alignment)"}
TPL_order <- fct_inorder(na.exclude(p$data$label[base::order(p$data$y, decreasing = TRUE)]), ordered = TRUE)

@@

Despite the presence of five *TPL/TPR* genes in the genome, which are considered to be redundant (pmid 16763149), very few observed SNPs are predicted to result in loss of function for any *TPL/TPR* gene. Only one frameshift variant was observed, *TPR4-665delG*, in a single accession (Aiell-4 from Italy). This deletion occurs in codon 222, just after the TPL N-terminal domain, and generates an early stop eleven codons downstream from the deletion. It is unclear whether this N-terminal fragment would be expressed at any appreciable level, as it is possible that the generation of an early stop codon would trigger nonsense-mediated decay. However, if expressed, it would be fascinating to examine whether this naturally occurring N-terminal truncation retains activity as a co-repressor. Given the observation that the TPL N-terminal domain is sufficient to bind histones (28630893), and repress transcription via synthetic assays in yeast (24979769), it is possible that this represents a semi-functional TPL variant. There has been no evidence for prioritization of individual *TPL/TPR* family members in the auxin response pathway, but experiments modulating the gene copy number of *TPR1* revealed the importance of maintaining the proper dosage of *TPL/TPR* in order to accurately trigger the salicylic acid innate immune response (20647385). These results, taken together with the absence of *TPL/TPR* loss-of-function alleles in the *Arabidopsis thaliana* accessions suggest that perhaps the entire family is being conserved at the nucleotide level in the population to meet a gene dosage requirement, with little opportunity for neo- or sub-functionalization.

>>>>

The TOPLESS co-repressor family generally exhibits a high level of sequence conservation at the amino acid sequence level across resequenced *Arabidopsis thaliana* accessions, with all ${\pi_N/\pi_S}$ values below 1 (Table X). The closely related *TPL* and *TPR1* have the highest ${\pi_N/\pi_S}$ values, suggesting that these these two related genes tolerate a higher degree of sequence and potentially functional diversity compared to *TPR2/3/4*. The N-terminal TPL domain of the TPL/TPR family is particularly conserved (Figure X). All nonsynonymous polymorphisms observed in this region are either in the coils between helices or are highly conservative mutations within helices (i.e. Valine to Isoleucine), which would be predicted to exhibit little effect on folding and function.

@@

@@

(ref:full-TPLs-alignment) **Alignment of A. thaliana TPL/TPR genes** 
```{r full-TPLs-alignment, fig.cap= "(ref:full-TPLs-alignment)"}
TPL_order <- fct_inorder(na.exclude(p$data$label[base::order(p$data$y, decreasing = TRUE)]), ordered = TRUE)

@@

### 2018-12-06 manuscript/index.Rmd
bioRxiv submission

---
title: Accelerating structure–function mapping using the ViVa webtool to mine natural variation 
author:  |
  | R. Clay Wright^[Corresponding author: wrightrc@vt.edu] $^1$, 
  | Morgan O. Hamm $^2$,
  | Jennifer Nemhauser $^2$
  | $^1$Biological Systems Engineering, Virginia Tech, Blacksburg, VA 24061, 
  | $^2$Deparment of Biology, University of Washington, Seattle, WA 98195

date: "`r Sys.Date()`"

abstract: Thousands of sequenced genomes are now publically available and represent a large swath of plant evolution; yet, much of this data remains inaccessible to researchers without significant bioinformatics experience. Here, we present a webtool called ViVa (Visualizing Variation) which aims to empower any researcher to take advantage of the amazing genetic resource collected in the Arabidopsis 1001 Genomes Project (http://1001genomes.org). ViVa facilates data mining on the gene, gene family or gene network level. To test the utility and accessibility of ViVa, we assembled a team with a range of expertise with biology and bioinformatics to analyze the natural variation in the well-studied nuclear auxin signaling pathway. Our analysis has provided further confirmation of existing knowledge and has also helped generate new hypotheses regarding this well studied pathway. These results highlight how natural variation could be used to generate and test hypotheses about less studied gene families and networks, especially when paired with biochemical and genetic characterization. ViVa is also readily extensible to newly emerging databases of genetic variation, such as the 3,000 Rice Genomes Project (http://snp-seek.irri.org/).

keywords: natural variation, Arabidopsis thaliana, genome diversity, structure, function, genotype, phenotype, bioconductor, bookdown, r markdown

bibliography:
- natural-variation.bib
- TIR1_AFB.bib
- Aux_IAA.bib
- ARF.bib
- TPL.bib
biblio-style: apalike
link-citations: yes

---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, collapse = TRUE, comment = "#>", tidy = TRUE, cache = TRUE, cache.lazy = FALSE, out.width = "90%", fig.height=4, fig.width=7, out.extra='style="margin: auto; display: block; padding-top: 15px;"')

#devtools::install_github("wrightrc/r1001genomes", ref = "auxin-natural-variation")
library(bookdown)
library(ape)
library(r1001genomes)
library(DT)
library(tidyverse)
library(magrittr)
library(RColorBrewer)
library(ggpmisc)
library(gginnards)
library(ggthemes)
library(ggseqlogo)
library(ggtree)
library(reticulate)
library(scales)
library(viridis)
library(ips)
library(knitr)
library(dplyr)
library(stringr)

@@

## List of working titles  
* Predicting structure--function relationships using natural genetic variation 
* Accelerating structure--function mapping using natural genetic variation 
* A comprehensive analysis of natural sequence variation within the <i>Arabidopsis thaliana</i> nuclear auxin signaling pathway 
* Visualizing natural variation with ViVa, a new webtool
* Accelerating structure–function mapping using a new webtool to mine natural variation

@@

>>>>

---
title: Accelerating structure-function mapping using the ViVa webtool to mine natural variation 
shorttitle: 'ViVa: Visualizing Variation' 
author:
  - Morgan O. Hamm^1^, Britney L. Moss^2^, Alexander R. Leydon^1^,  Hardik P. Gala^1^,  Amy Lanctot^1^, Romi Ramos^1^, Hannah Klaeser^2^, Andrew C. Lemmex^1^, Mollye L. Zahler^1^, Jennifer L. Nemhauser^1^, R. Clay Wright^3^$*$ 
  - ^1^Department of Biology, University of Washington, Seattle, WA, USA.
  - ^2^Department of Biology, Whitman College, Walla Walla, WA, USA. 
  - ^3^Biological Systems Engineering, Virginia Tech, Blacksburg, VA, USA. Previous affiliation$:$ Department of Biology, University of Washington, Seattle, WA, USA.
  - $*$Direct correspondence to$:$ wrightrc@vt.edu.

date: "`r Sys.Date()`"
abstract: "Thousands of sequenced genomes are now publicly available capturing a significant amount of natural variation within plant species; yet, much of this data remains inaccessible to researchers without significant bioinformatics experience. Here, we present a webtool called ViVa (Visualizing Variation) which aims to empower any researcher to take advantage of the amazing genetic resource collected in the *Arabidopsis thaliana* 1001 Genomes Project (http://1001genomes.org). ViVa facilitates data mining on the gene, gene family or gene network level. To test the utility and accessibility of ViVa, we assembled a team with a range of expertise within biology and bioinformatics to analyze the natural variation within the well-studied nuclear auxin signaling pathway. Our analysis has provided further confirmation of existing knowledge and has also helped generate new hypotheses regarding this well studied pathway. These results highlight how natural variation could be used to generate and test hypotheses about less studied gene families and networks, especially when paired with biochemical and genetic characterization. ViVa is also readily extensible to databases of interspecific genetic variation in plants as well as other organisms, such as the 3,000 Rice Genomes Project (http://snp-seek.irri.org/) and human genetic variation (https://www.ncbi.nlm.nih.gov/clinvar/).

Keywords: natural variation, Arabidopsis thaliana, genome diversity, structure, function, genotype, phenotype, bioconductor, bookdown, r

"

bibliography: 
- natural-variation.bib 
- packages.bib

biblio-style: apalike
link-citations: yes
fig_caption: yes
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE, collapse = TRUE, comment = "#>", tidy = TRUE, cache = TRUE, cache.lazy = FALSE, fig.height=4, fig.width=7, dpi = 600, out.extra='style="margin: auto; display: block; padding-top: 15px;"')

#devtools::install_github("wrightrc/r1001genomes", ref = "auxin-natural-variation")
library(bookdown)
library(ape)
library(r1001genomes)
library(DT)
library(tidyverse)
library(magrittr)
library(RColorBrewer)
library(ggpmisc)
library(gginnards)
library(ggthemes)
library(ggseqlogo)
library(ggtree)
library(reticulate)
library(scales)
library(viridis)
library(ips)
library(knitr)
library(dplyr)
library(stringr)
library(treeio)

@@

@@

\newpage
